\propertitle

### 2019-07-09 maxNt_v5.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-08T23:05:19+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

In the neurosciences, the maximum-entropy method has often been applied to
questions of sufficient statistics. In this section we discuss how our
proposed application bears on this kind of questions. For concreteness
we use an example with specific statistics: first and second moments
(equivalent to mean and correlations), versus first-to-fourth moments. But
our discussion holds for any other sets of statistics.

The question is as follows. We want to assess our degree of belief about
the frequencies of the activities of the sample, or about the sample
activity in a new time bin. For this assessment we use any measured data we
have. But it sometimes happens that dropping part of the data -- for
example, the measured third- and higher-order moments -- leaves our
assessment almost unchanged: the dropped data are \emph{informationally
  irrelevant}, or \emph{almost} so. The data that remain -- for example,
first and second moments -- are \emph{informationally sufficient}. This
informational quasi-sufficiency may be connected to biological or physical
mechanisms; that's why it's interesting.

We can approximately check which data are irrelevant by simply dropping
them from the conditional of our degree of belief, and see if the latter
changes appreciably. When our degree of belief is built with the
maximum-entropy method, we simply drop some of the data used as constraints
and see how the resulting distribution changes. In the neurosciences this
has been done with maximum-entropy distributions constructed \emph{for the
  sample}.

@@

The method introduced in the present work reveals an issue with this
application to the sample. The issue doesn't come from the method itself
but from a simple, general fact of probability theory. If a probability
distribution has some sufficient statistics
\citep[\sect~4.5]{bernardoetal1994}{fortinietal2000}, then its marginals
\emph{cannot} have the same sufficient statistics, and vice versa, except
for trivial cases such as uniform distributions. This in general also holds
between different marginals: if a marginal has some sufficient statistics,
another won't have it. This impossibility is known in statistical
mechanics: marginals of Gibbs states aren't Gibbs states
\citep{maesetal1999}. Mathematically it corresponds to the general
impossibility of solving a system of $n$ equations in $m$ unknowns when
$n>m$ \citep[\sect~3.1]{portamanaetal2015}[\sect~3]{portamanaetal2018b}.

In the neuroscientific application above, this means that if pairwise
correlations, for example, are sufficient statistics for a particular
sample of from a brain area, then they \emph{cannot} be sufficient for the
full network of neurons in that area, or for a different sample; vice
versa, if we assume that they are sufficient statistics for the full
network, then we must expect them \emph{not} to be sufficient for any
recorded sample thereof. But questions about sufficient statistics are
meant to be addressed to a whole brain area, not just to some casually
recorded sample. Thus, maximum-entropy methods applied at the sample level
paradoxically give an answer opposite to the one we might think they're
giving.

Questions about sufficient statistics are therefore correctly addressed by
applying the maximum-entropy method at the full-network level. If a
comparison with the measured frequencies of the sample activities is
desired, then the marginal distribution to the sample size should be used.

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/comparison_full.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/comparison_marginals.pdf}%
\caption{***}
\label{fig:comparisons_marginals}
\end{figure}
%comparison_ME_sample_full_levels.nb
Does the application to the full-network lead to very different results
from the one to the sample? Figure~\ref{fig:comparisons_marginals} gives a
graphical answer in the case of the first data set. The upper plot shows
the maximum-entropy distributions for a full network of $10\,000$ neurons,
constructed from two moments (\textcolor{mygreen}{dashed green curve}) and
from four moments (\textcolor{myblue}{solid blue curve}). The lower plot
shows the maximum-entropy distributions for the sample (empty markers),
from two moments (\textcolor{mygreen}{empty green triangles}) and from four
moments (\textcolor{myblue}{empty blue squares}), and the distributions
obtained by marginalizing the full-network distribution to the sample size
(filled markers), from two moments (\textcolor{mygreen}{filled green
  triangles}) and from four moments (\textcolor{myblue}{filled blue
  squares}). The measured frequency distribution (\textcolor{myred}{red
  circles}) is also shown. We note the following:
\begin{itemize}
\item The application to the full-network (upper plot) leads to two
  completely different distributions (even different number of modes);
  clearly the two statistics are not even approximately equivalent for
  inferential purposes.
\item The application to the sample leads to deceivingly similar
  distributions (empty triangles and empty squares), which can only be
  distinguished on a logarithmic scale. This could lead to the erroneous
  conclusion that the two statistics are approximately equivalent.
\item The difference between marginals from the full-network distribution
  and the distributions obtained at the sample level (filled vs empty
  markers) is larger than the difference between different statistics
  (triangles vs squares).
\item The marginals of the application to the full network are closer to
  the measured frequencies than the distributions obtained at the sample
  level (although this closeness is not a valid criterion for their
  goodness).
\end{itemize}
Therefore, the two maximum-entropy applications lead to different results,
and the application at the full-network level is clearly superior, leading
to a clearer distinction between the effect of considering different
statistics.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-09T13:33:41+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

In the neurosciences the maximum-entropy method is often applied to
questions of sufficient statistics. In this section we discuss how our
proposed application bears on this kind of questions. For concreteness we
use an example with specific statistics: first and second moments
(equivalent to mean and correlations), versus first-to-fourth moments. But
our discussion holds for any  sets of statistics.

The question is as follows. We want to assess our degree of belief about
the frequencies of the activities of the sample, or about the sample
activity in a new time bin. For this assessment we should use all measured
data we have. But it sometimes happens that dropping part of the data --
for example, the measured third- and higher-order moments -- leaves our
assessment almost unchanged: the dropped data are \emph{informationally
  irrelevant}, or almost so. The remaining data -- for example, first and
second moments -- are \emph{informationally sufficient}. This informational
quasi-sufficiency is interesting because it may hint at peculiar biological
or physical mechanisms or dependencies.

We can approximately check how much some data are irrelevant by simply
dropping them from the conditional of our degree of belief, and see if the
latter changes appreciably. When our degree of belief is built with the
maximum-entropy method, we simply drop some of the data used as constraints
and see how the resulting distribution changes. For example, we may only
retain the first two moments (means and correlations) or the first four
moments. In the neurosciences this check is often done with maximum-entropy
distributions constructed \emph{for the sample}, ignoring the full network
it's sampled from.

@@

The method introduced in the present work reveals an issue with this
application at the sample level. The issue doesn't come from the method
itself but from a general fact of probability theory. If a probability
distribution has some sufficient statistics
\citep[\sect~4.5]{bernardoetal1994}{fortinietal2000}, then its marginals
\emph{cannot} have the same sufficient statistics, and vice versa, except
for trivial cases such as uniform distributions. This generally also holds
among different marginals: if a marginal has some sufficient statistics,
another won't have it. This impossibility is known in statistical
mechanics: marginals of Gibbs states aren't Gibbs states
\citep{maesetal1999}. Mathematically it corresponds to the general
impossibility of solving a system of $n$ equations in $m$ unknowns when
$n>m$ \citep[\sect~3.1]{portamanaetal2015}[\sect~3]{portamanaetal2018b}.

In our example this means that if means and pairwise correlations or the
first four moments are informationally sufficient for a particular sample
from a brain area, then they \emph{cannot} be sufficient for the full
network of neurons in that area, or for a different sample. Vice versa, if
we assume that they are informationally sufficient for the full network,
then we must expect them \emph{not} to be sufficient for any recorded
sample thereof. Now, questions about sufficient statistics are meant to be
addressed to a whole brain area, not just to some casually recorded sample.
Thus, paradoxically, maximum-entropy methods applied at the sample level
give an answer opposite to the one we might think they're giving.

Questions about informational sufficiency are therefore correctly addressed
by applying the maximum-entropy method at the full-network level. If a
comparison with the measured frequencies of the sample activities is
desired, then the marginal distribution to the sample size should be used.

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/comparison_full.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/comparison_marginals.pdf}%
\caption{***}
\label{fig:comparisons_marginals}
\end{figure}
%comparison_ME_sample_full_levels.nb
But does the application at the full-network level lead to appreciably
different results from that at the sample level? after all we're interested
in an approximate informational sufficiency, not in an analytically exact
one. Figure~\ref{fig:comparisons_marginals} gives a graphical answer to
this question in the case of our first data set. The upper plot shows the
maximum-entropy distributions for a full network of $10\,000$ neurons,
constructed from two moments (\textcolor{mygreen}{\textbf{--~--~--}}) and
from four moments (\textcolor{myblue}{\textbf{------}}). The lower plot
shows the maximum-entropy distributions for the sample, from two moments
(\textcolor{mygreen}{$\bm{\triangledown}$}) and from four moments
(\textcolor{myblue}{$\bm{\square}$}), and the distributions obtained by
marginalizing the full-network distribution to the sample size, from two moments (\textcolor{mygreen}{$\bm{\blacktriangledown}$})
and from four moments (\textcolor{myblue}{$\bm{\blacksquare}$}). The
measured frequency distribution (\textcolor{myred}{\LARGE$\bm{\bullet}$}) is also
shown. We note the following:
\begin{itemize}
\item The application to the full-network (upper plot) leads to two
  completely different distributions (even different number of modes);
  clearly the two sets of statistics are not even approximately equivalent
  for inferential purposes.
\item The application at the sample level leads to deceivingly similar
  distributions (\textcolor{mygreen}{$\bm{\triangledown}$} and
  \textcolor{myblue}{$\bm{\square}$}), which can only be distinguished on a
  logarithmic scale. This could lead to the erroneous conclusion that the
  two statistics are approximately equivalent.
\item The difference between marginals from the full-network distribution
  and the distributions obtained at the sample level (filled vs empty
  markers) is larger than the difference between different statistics
  (triangles vs squares).
\item The marginals of the application to the full network are closer to
  the measured frequencies than the distributions obtained at the sample
  level (although this closeness is not a valid criterion for their
  goodness).
\end{itemize}
Therefore, the maximum-entropy application at the sample level and at the
full-network level lead to different results; the latter is not only more
meaningful but also superior because it shows clearly the different
informational sufficiency of the two sets of statistics.
